\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2024-05-25 doc/paper-jss/sdmTMB-paper.Rnw
Revisions - residuals - BLAS - convergence changes - Moran's I - statsv typo

\Abstract{
Geostatistical spatial or spatiotemporal data are common across scientific fields.
% and the dimensions of these data continue to grow in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian Markov random fields (GMRFs), are computationally intensive and challenging for many users to implement.
Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent GMRFs using an SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \pkg{fmesher} and estimation is conducted via maximum marginal likelihood with \TMB\ or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), hurdle models, cross-validation and anisotropy (directionally dependent spatial correlation).
% Additional functionality includes break-point effects, out-of-sample cross validation, and priors.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
We hope \sdmTMB's accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

\subsubsection{Traditional delta models}

@@

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} as interface to the \pkg{Eigen} library in C++.
% TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023). _Matrix: Sparse and Dense Matrix Classes and Methods_. R package version 1.6-4, <https://CRAN.R-project.org/package=Matrix>.)), as central dependency of TMB, given the detailed description of the dependency stack. If really motivated, you could also add the reference for Eigen (Guennebaud G, Jacob B, et al. (2010). “Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{statsv::optimHess} \citep{gay1990, r2021}.
\Rev{A19}Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
it is also possible to apply the epsilon estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness of random effects when calculating an estimator as a non-linear transformation or random effects (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

@@

A variety of model evaluation tools are available.
A \fct{residuals} method calculates various types of residuals.
New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
\fct{sdmTMB\_cv} facilitates cross validation.

@@

@@

We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \fct{residuals} method.
We omit example code for Moran's I calculation for brevity but include it in the code supplement.
We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but no evidence for the model with a random field (p = 0.84) suggesting that residual spatial autocorrelation is alleviated by including the random field.

<<morans, echo=FALSE, results='hide'>>=
test_autocor <- function(obj) {
  set.seed(1)
  s <- simulate(obj, nsim = 500)
  pr <- predict(obj, type = "response")$est
  r <- DHARMa::createDHARMa(
    simulatedResponse = s,
    observedResponse = pcod$present,
    fittedPredictedResponse = pr
  )
  DHARMa::testSpatialAutocorrelation(r, x = pcod$X, y = pcod$Y, plot = FALSE)
}
(t_no_rf <- test_autocor(fit_bin))
(t_rf <- test_autocor(fit_bin_rf))
p_rf <- round(t_rf$p.value, 2)
@

@@

Indeed, the log likelihood of left-out data is considerably larger for the model that includes random fields indicating better out-of-sample predictions:

@@

In practice, we would want to repeat this procedure several times to ensure the rank order is not sensitive to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.

@@

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta model formulation as described in Section \ref{sec:delta}.
Note that we choose to omit the spatiotemporal random fields from the binomial model component (\code{spatiotemporal = list("off", "iid")}) because initial experimentation indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.

<<check-version, results='hide', echo=FALSE, include=FALSE>>=
# delta_poisson_link_gamma() deprecated in favor of delta_gamma(type = "poisson-link")
new_deltas <- packageVersion("sdmTMB") > '0.4.2.9000'
@

<<dog-update, results='hide', message=FALSE, warning=FALSE, echo=new_deltas, eval=new_deltas, include=new_deltas>>=
fit_dg <- update(fit_tw, family = delta_gamma(),
  spatiotemporal = list("off", "iid"))
fit_dl <- update(fit_dg, family = delta_lognormal())
fit_dpg <- update(fit_dg, family = delta_gamma(type = "poisson-link"))
fit_dpl <- update(fit_dg, family = delta_lognormal(type = "poisson-link"))
@

<<dog-update-old, results='hide', message=FALSE, warning=FALSE, echo=!new_deltas, eval=!new_deltas, include=!new_deltas>>=
fit_dg <- update(fit_tw, family = delta_gamma(),
  spatiotemporal = list("off", "iid"))
fit_dl <- update(fit_dg, family = delta_lognormal())
fit_dpg <- update(fit_dg, family = delta_poisson_link_gamma())
fit_dpl <- update(fit_dg, family = delta_poisson_link_lognormal())
@

@@

We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the positive model component spatiotemporal random fields as AR(1) to allow spatiotemporal hotspots to persist from year to year.

@@

<<dog-ar1, results='hide', echo=TRUE>>=
fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = list("off", "ar1"))
@

@@

<<dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}. Ellipses are centered at coordinates of zero in the units that the X-Y coordinates are modeled. The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from the center (zero).", out.width="4in">>=
plot_anisotropy(fit_dpl)
@

On inspecting the model output with \fct{summary}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
We therefore consider a model where these are turned off in the second component of the delta model and choose this model for further evaluation.

<<dog-ar1-2, echo=TRUE, results="hide">>=
fit_dpl_ar1_only <- update(fit_dpl_ar1, spatial = list("on", "off"))
@

<<dog-aic3, echo=TRUE>>=
AIC(fit_dpl_ar1_only, fit_dpl_ar1, fit_dpl)
@

@@

<<dog-print, echo=TRUE>>=
fit <- fit_dpl_ar1_only
sanity(fit)
summary(fit)
@

@@

@@

\input{response}

>>>>

\Abstract{
Geostatistical spatial or spatiotemporal data are common across scientific fields.
% and the dimensions of these data continue to grow in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian Markov random fields (GMRFs), are computationally intensive and challenging for many users to implement.
Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent GMRFs using an SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \pkg{fmesher} and estimation is conducted via maximum marginal likelihood with \TMB\ or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), hurdle models, cross-validation and anisotropy (directionally dependent spatial correlation).
% Additional functionality includes break-point effects, out-of-sample cross validation, and priors.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
We hope \sdmTMB will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

\subsubsection{Standard delta models}

@@

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} as interface to the \pkg{Eigen} library in \proglang{C++}.
% TODO Jim:  I recommend adding a citation to package Matrix ( Bates D, Maechler M, Jagan M (2023). _Matrix: Sparse and Dense Matrix Classes and Methods_. R package version 1.6-4, <https://CRAN.R-project.org/package=Matrix>.)), as central dependency of TMB, given the detailed description of the dependency stack. If really motivated, you could also add the reference for Eigen (Guennebaud G, Jacob B, et al. (2010). “Eigen v3.” http://eigen.tuxfamily.org), although it's less clear what citation to use.
The negative marginal log likelihood ivs minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2021}.
\Rev{A19}Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness caused by non-linear transformations of random effects (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

@@

A variety of model evaluation tools are available.
A \fct{residuals} method calculates various types of residuals.
The default is a form of randomized quantile residuals \citep{dunn1996}.
New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
\fct{sdmTMB\_cv} facilitates cross validation.

@@

Users willing to replace the default \R\ \pkg{BLAS} \citep{blas}
% and \pkg{LAPACK} \citep{lapack}
library
with an optimized version \citep[e.g., \pkg{openBLAS};][]{openblas}
can expect about an order of magnitude increase in model fitting speed.
Suggestions are included in the package README.

@@

We can test for spatial autocorrelation with a visual inspection or a statistical test of the residuals.
Here, we demonstrate an approach using an implementation of Moran's I from the \pkg{ape} package \citep{gittleman1990, ape}.
We set \code{type = "mle-mvn"} to denote setting fixed effects at their maximum likelihood estimates but taking a single draw from the approximate (multivariate normal) distribution of the random effects \citep{waagepetersen2006, thygesen2017a}.

<<morans, echo=TRUE>>=
inv_dist_matrix <- 1 / as.matrix(dist(pcod[,c("X", "Y"), ]))
diag(inv_dist_matrix) <- 0
set.seed(1)
r_bin <- residuals(fit_bin, type = "mle-mvn")
set.seed(1)
r_bin_rf <- residuals(fit_bin_rf, type = "mle-mvn")

ape::Moran.I(r_bin, weight = inv_dist_matrix)$p.value
ape::Moran.I(r_bin_rf, weight = inv_dist_matrix)$p.value
@

% We omit example code for Moran's I calculation for brevity but include it in the code supplement.
We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for the model with a random field (p $>$ 0.01) suggesting that residual spatial autocorrelation is alleviated by including the random field.
We note that the specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals \citep{dunn1996}.

@@

Indeed, the log likelihood of left-out data is considerably higher for the model that includes random fields indicating better out-of-sample predictions:

@@

In practice, we would repeat this procedure several times to ensure the rank order is not sensitive to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.

@@

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta model formulation as described in Section \ref{sec:delta}.
Note that we need to add an extra Newton optimization step to the delta-lognormal model to achieve suitable convergence (discovered via a call to \code{sanity(fit\_dl)}).
% Note that we choose to omit the spatiotemporal random fields from the binomial model component (\code{spatiotemporal = list("off", "iid")}) because initial experimentation indicated this random field was collapsing to zero indicating minimal changes in encounter spatial distribution from year to year.

<<dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE>>=
fit_dg <- update(fit_tw, family = delta_gamma())
fit_dl <- update(fit_tw, family = delta_lognormal(),
  control = sdmTMBcontrol(newton_loops = 2))
fit_dpg <- update(fit_tw, family = delta_gamma(type = "poisson-link"))
fit_dpl <- update(fit_tw, family = delta_lognormal(type = "poisson-link"))
@

@@

We next test two additional model formulations: making the spatial correlation isotropic (the default) instead of anisotropic, and structuring the positive model component spatiotemporal random fields as AR(1) to allow spatiotemporal patterns to partially persist from year to year.

@@

<<dog-ar1, results='hide', echo=TRUE>>=
fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = "ar1")
@

@@

<<dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}. Ellipses are centered at coordinates of zero in the units that the X-Y coordinates are modeled. The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from the center (zero).", out.width="4in">>=
plot_anisotropy(fit_dpl_ar1)
@

@@

<<dog-print, echo=TRUE>>=
fit <- fit_dpl_ar1
sanity(fit)
summary(fit)
@

@@

We can check simulation-based randomized quantile residuals from our chosen model via the \pkg{DHARMa} package \citep{dharma}.
To do that, we simulate from our model with the \fct{simulate.sdmTMB} method and pass those simulations to a helper function \fct{dharma\_residuals}.

<<dog-residuals, fig.cap="Simulation-based randomized quantile residuals from the \\pkg{DHARMa} package. The statistical test is a two-sided Kolmogorov-Smirnov test of the null hypothesis that the residuals are drawn from a uniform(0, 1) distribution.", fig.asp=0.68, echo=TRUE>>=
set.seed(1)
s <- simulate(fit, nsim = 500, type = "mle-mvn")
dharma_residuals(s, fit)
@

The quantile-quantile plot suggests that under the model assumptions, the (transformed) residuals are consistent with an independent uniform(0, 1) distribution.

@@

% \input{../../../sdmTMB-paper-response/response}
